Rationality, Evolution \& the Environment

### 2015-06-05 paper/paper.tex
Paolo rewrite Version 1 of Section 2

  \citep[e.g.][]{ZollmanSmead2010:Plasticity-and-,SmeadZollman2013:The-Stability-o}. (\myalert{More
    references Harley 1981!!!, \dots}). Moreover, only little attention has been paid to

@@

A pivotal element of economics and decision theory is the definition of rationality as
maximization of subjective preferences. The subjectivity of preferences plays a very important
role. From an agent's standpoint, the amount of, say, money achieved does not necessarily
coincide with the amount of subjective utility attained. The first to notice this difference
between objective utility (goods, money, etc.) and subjective utility (preferences) was
D. Bernoulli in the famous St. Petersburg paradox. Moreover, Bernoulli introduced the idea that
subjective utility can be represented by an increasing concave function of objective
utility. This can be considered the birth of modern decision theory.

Another important step was developed by Kahnemann and Tversky who introduced prospect theory in
their famous paper from 1979. By means of experimental results they noticed that the decision
maker apparently acts as a risk averse agent with respect to gains and as a risk loving agent
with respect to losses. Consequently, they conjectured that the subjective utility is better
representable by an S-shaped utility function, concave in the positive domain of gains and
convex in the negative domain of losses.

More recently, other possible definitions of the subjective utility function have been proposed
and justified from a game theoretical point of view (e.g. Fehr and Schmidt, Charness and Rabin,
etc.). The central goal of these studies is to explain/describe the behavior in social dilemmas
(e.g., Prisoner's dilemma, public goods game, etc.). By means of different subjective utility
functions, it is then possible to describe/match the behavior observed in experiments and to
explain it as rational (in the sense of maximizing the subjective utility).

With this approach it seems that every behavior, no matter how far from the maximization of
objective utility, can be deemed as rational if we can only find a subjective utility that
justifies that behavior as maximizing the subjective utility. Quoting Hegel, (almost)
everything real is rational. The problem with this approach is that it seems that everything
that is observed can be considered rational, given the adequate subjective utility. The notion
of rationality almost collapses onto the observed behavior.

On the other side, other economists (Alchian, Friedman) seem to link the notion of rationality
more tightly to the maximization of objective utility. In their classic works they argue that
profit maximization is a reasonable assumption for characterizing outcomes in a competitive
market because only firms behaving in a manner that is consistent with profit maximization will
survive in the long run. If the previous accounts of rationality looked more descriptive, this
definitely tastes more normative. In this sense, the normative notion of rationality is tightly
linked to (and almost coincide with) the maximization of objective utility, which is supposed
to drive the evolutionary process. It follows then that being normative in this sense seems to
entail an identification of rationality with maximization of objective utility.

It seems that we are stuck between two extrema. On the one hand, if we want to have a
descriptive notion of rationality, then it seems we can (almost) always define an appropriate
subjective utility whose maximization matches the observed behavior. On the other hand, if we
want to have a normative approach to rationality, then it seems we have to appeal to
evolutionary arguments implying the understanding of rationality as maximization of objective
utility.

The meta-game perspective that we introduce here can shed more light on this duality, and opens
some space in between the two extrema. By keeping the evolutionary approach we want to retain a
normative perspective on rationality, but our results show that (under certain conditions)
evolution does not necessarily lead to identifying rationality with maximization of objective
utility. More precisely, from the meta-game perspective many possible subjective
transformations of the objective utility are eliminated by evolution, but we find some
evolutionarily stable subjective utility maximizers whose utility does not coincide with the
objective evolutionary utility. This result leaves some room in between the two extrema: we can
understand rationality in a normative way, given by evolution, without having to discard all
the possible subjective utility representations as irrational.

In particular, in the following we mainly deal with two types of agents: players whose utility
function coincides with the objective utility, and players whose subjective utility is defined
in terms of regret. The justification for this choice is both historical and behavioral. Among
the possible competitors of utility maximization, regret has always been in a prominent
position, both from a descriptive and normative viewpoint. Indeed, Bell82 writes
\begin{quote}
  By explicitly incorporating regret, expected utility theory not only becomes a better
  descriptive predictor but also may become a more convincing guide for prescribing behavior to
  decision makers.
\end{quote}
\citet{LoomesSugden1982:Regret-Theory:-} show that a regret-based theory of choice can match
and explain the main experimental results of prospect theory, and claim that
\begin{quote} [...] we shall challenge the idea that the conventional axioms [of decision
  theory] constitute the only acceptable basis for rational choice under uncertainty. We shall
  argue that it is no less rational to act in accordance with regret theory, and that
  conventional expected utility theory therefore represents an unnecessarily restrictive notion
  of rationality. \hfill \citep{LoomesSugden1982:Regret-Theory:-}
\end{quote}

@@

>>>>

  \citep[e.g.][]{ZollmanSmead2010:Plasticity-and-,SmeadZollman2013:The-Stability-o}. (\myalert{More references Harley 1981!!!, \dots}). 
    Moreover, only little attention has been paid to

@@

The standard use of the word \textit{rationality} in economics and decision theory is to denote the maximization of subjective expected utility. Since the work by vNM [cite], this stance has been justified from the intuitive ''rationality'' of the axioms on the subjective preference relation leading to the representability of the decision maker as a subjective expected utility (SEU) maximizer. 
While in [vNM] the probability of the outcomes is objectively given a priori, Savage54 provides a set of axioms that represent the decision maker (DM) with both a subjective utility over outcomes \textit{and} a \textit{subjective} probability over states of the world. Savage's axioms constitute the foundation of subjective expected utility theory, that is normally assumed to define the rational behavior of economic agents.

The subjectivity of preferences plays a very important
role. From an agent's standpoint, the amount of, say, money achieved does not necessarily
coincide with the amount of subjective utility attained. The first to notice and formally argue for this difference
between objective utility (goods, money, etc.) and subjective utility (preferences) was
D. Bernoulli in the famous St. Petersburg paradox. Moreover, Bernoulli introduced the idea that a good representation of
subjective utility can be carried out in terms of an increasing concave function of objective
utility. This can be considered the birth of modern decision theory.

However, many violations of Savage's (and vNM's) axioms have been consistently observed in behavioral economics. Most of the attention has been focused on the Independence Axiom and the Sure-thing Principle. Allais' paradox and Ellsberg's paradox are examples of violations of these two axioms, and they are at the origin of Prospect theory [cite] and ambiguity in decision theory [cite].\\

The main idea behind Ellesberg's paradox and the literature on ambiguity is that DM treats differently objective and subjective uncertainty. If the probability measure on the possible states of the world is not given a priori and ''known'', then observed preferences leave no room to represent  DM's beliefs via a probability measure on the possible states. The most common solution is then to represent DM's beliefs with a set of probability measures instead of a single one.

Prospect theory was introduced by Kahnemann and Tversky in
their famous paper from 1979. By means of experimental results they noticed that the decision
maker apparently acts as a risk averse agent with respect to gains and as a risk loving agent
with respect to losses. Consequently, they conjectured that the subjective utility is better
representable by an S-shaped utility function, concave in the positive domain of gains and
convex in the negative domain of losses.
\\

More recently, other possible definitions of the subjective utility function have been proposed
and justified from a game theoretical point of view (e.g. Fehr and Schmidt, Charness and Rabin,
etc.). The central goal of these studies is to explain/describe the behavior in social dilemmas
(e.g., Prisoner's dilemma, public goods game, etc.). In the same spirit of Prospect theory, by means of different subjective utility
functions it is possible to describe/match the behavior observed in game theoretical experiments and to
explain it as rational (in the sense of maximizing the subjective utility).

With this approach every behavior, no matter how far from the maximization of
objective utility, can be deemed rational if we can only find a subjective utility that
justifies that behavior as maximizing the subjective utility. Quoting Hegel, (almost)
everything real is rational (this is just for fun, I will take it away, no problem). The problem here is that it seems possible to consider rational everything
that is observed, given the adequate subjective utility. The notion
of rationality almost collapses onto the observed behavior.

On the other side, other economists (Alchian, Friedman) seem to link the notion of rational choice
more tightly to the maximization of objective utility. In their classic works they argue that
profit maximization is a reasonable assumption for characterizing outcomes in a competitive
market because only firms behaving in a manner that is consistent with profit maximization will
survive in the long run. If the previous accounts of rationality looked more descriptive, this
definitely tastes more normative. In this sense, the normative notion of rationality is tightly
linked to (and almost coincide with) the maximization of objective utility, which is supposed
to drive the evolutionary process. It follows then that being normative in this sense points to an identification of rationality with maximization of objective utility.

It seems that we are stuck between two extrema: a too permissive descriptivism and a too restrictive normativism. On the one hand, if we want to have a
descriptive notion of rationality, then we can (almost) always define an appropriate
subjective utility whose maximization matches the observed behavior. On the other hand, if we
want to have a normative approach to rationality, then it seems we have to appeal to
evolutionary arguments entailing the understanding of rationality as maximization of objective
utility.

There is another violation of Savage's axioms that is particularly relevant for the present work: transitivity of preferences. The first of Savage's axioms states that the preference relation is a (total) preorder, i.e., reflexive and transitive. Moreover, the preorder axiom is often referred to as the rationality axiom [cite], as to suggest that it is somehow the core of rational preferences. Violations of transitivity have been observed and studied in experimental literature [citeTversky69,etc], but can also derive from a more theoretical and normative point of view. \\
In what follows, an important focus of our work will be the role of regret in decision making. Regret theory predicts the violation of both transitivity and Independence of Irrelevant Alternatives (IIA). [CONNECTING TO Fawcett\&al: this is IIA, not independence]

\subsubsection{Regret theory}

Regret theory has been introduced in decision theory by Savage [] and Niherens [], and later developed by Bell [1982], Loomes and Sugden [1982], and Fishburn [1982] independently. Lately, Halpern and Pass show how the use of regret minimization can solve games like the Centipede and the Traveler's dilemma in a way closer to everyday intuition and empirical data. In this paper the notion of regret is defined as in [HalPass12]:

DEFINITION(regret): Given a simultaneous-move game $ G=\langle N, (A_i , \pi_i)_{i \in N} \rangle $, the regret of an action $a_i \in A_i$ is defined by: $\text{reg}(a_i)= \text{max}_{a_{-i}\in A_{-i}} \pi_i(a_i^*,a_{-i})-\pi_i(a_i,a_{-i}) $, where $a_i^*$ is $i$'s best reply to action $a_{-i} $.\\

DEFINITION(regret minimization): Given a simultaneous-move game $ G=\langle N, (A_i , \pi_i)_{i \in N} \rangle $, an action $a^{'}_i $ is regret minimizing if $a^{'}_i \in \text{min}_{a_i} \text{max}_{a_{-i}} \pi_i(a_i^*,a_{-i})-\pi_i(a_i,a_{-i}) $.\\

EXAMPLE. Traveler's dilemma maybe? or maybe 2

The meta-game perspective that we introduce here can shed more light on the duality of descriptivism and normativism, and opens
some space in between the two extrema. 
By keeping the evolutionary approach we want to retain a
normative perspective on rationality, but our results show that (under certain conditions)
evolution does not necessarily lead to identifying rationality with maximization of objective
utility.

[More precisely, from the meta-game perspective many possible subjective
transformations of the objective utility are eliminated by evolution, but we find some
evolutionarily stable subjective utility maximizers whose utility does not coincide with the
objective evolutionary utility. This result leaves some room in between the two extrema: we can
understand rationality in a normative way, given by evolution, without having to discard all
the possible subjective utility representations as irrational. CHANGE PLACE OF THIS MAYBE]

In the following we mainly deal with two types of agents: players whose subjective utility is defined
in terms of regret (as in DEFINITION), and players whose utility
function coincides with the objective utility $\pi_i$.

The justification for this choice is both historical and behavioral. Among
the possible competitors of utility maximization, regret has always been in a prominent
position, both from a descriptive and normative viewpoint. Indeed, Bell82 writes
\begin{quote}
  By explicitly incorporating regret, expected utility theory not only becomes a better
  descriptive predictor but also may become a more convincing guide for prescribing behavior to
  decision makers.
\end{quote}
\citet{LoomesSugden1982:Regret-Theory:-} show that a regret-based theory of choice can match
and explain all the main experimental results of prospect theory, and claim that
\begin{quote} [...] we shall challenge the idea that the conventional axioms [of decision
  theory] constitute the only acceptable basis for rational choice under uncertainty. We shall
  argue that it is no less rational to act in accordance with regret theory, and that
  conventional expected utility theory therefore represents an unnecessarily restrictive notion
  of rationality. \hfill \citep{LoomesSugden1982:Regret-Theory:-}
\end{quote}

@@

[A pivotal element of economics and decision theory is the definition of rationality as
maximization of subjective preferences.]
